Catching runtime errors of Redis at compile time

### 2015-11-15 paper.tex
introduction

\begin{abstract}
We often interacts with database systems through dedicated libraries or packages
implemented in some programming languages. People make mistakes, while some of
the syntactic and semantic errors can be prevented by the language at compile
time, or caught by the package at runtime, most semantic errors are just
being ignored, causing problems at the database system.

In this paper, we demonstrates how to prevent those runtime errors at compile
time, and allow users to write reliable database queries without runtime overhead,
by exploiting type-level programming techniques such as parametedzied monad,
type-level literals and type families in Haskell.

@@

The previous example in Redis's client would look something like this with \emph{Hedis}
in Haskell.
\begin{minted}{haskell}
program :: Redis (Either Reply [ByteString])
program = do
    sadd "some-set" ["a", "b"]
    sadd "another-set" ["a", "b", "C"]
    sinter ["some-set", "another-set"]
\end{minted}

Function \mintinline{haskell}{sadd} takes a key and a list of values as arguments
, and returns an \mintinline{haskell}{Integer}, wrapped in
\mintinline{haskell}{Either} to indicate possible failures in
the context \mintinline{haskell}{Redis} of command execution.

@@

Keys and values are of type \mintinline{haskell}{ByteString}, since they are all
just binary strings in Redis. That means if a user wants to store values of
arbitrary types, he or she will have to encode and decode them as
 \mintinline{haskell}{ByteString}s.

>>>>

\begin{abstract}
Programmers often interact with database systems, by sending queries through
libraries or packages in some programming languages. But people make mistakes,
while some of the syntactic and semantic errors can be prevented by the language
at compile time, or caught by the package at runtime, most semantic errors are
just being ignored, causing problems at the database system.

In this paper, we demonstrates how to prevent those runtime errors at compile
time, allowing users to write reliable database queries without runtime
overhead, by exploiting type-level programming techniques such as parametedzied
monad, type-level literals and type families in Haskell.

@@

The previous example in Redis's client would look something like this with
\emph{Hedis} in Haskell.
\begin{minted}{haskell}
program :: Redis (Either Reply [ByteString])
program = do
    sadd "some-set" ["a", "b"]
    sadd "another-set" ["a", "b", "c"]
    sinter ["some-set", "another-set"]
\end{minted}

Function \mintinline{haskell}{sadd} takes a key and a list of values as
arguments, and returns an \mintinline{haskell}{Integer}, wrapped in
\mintinline{haskell}{Either} to indicate possible failures in the context
\mintinline{haskell}{Redis} of command execution.

@@

Keys and values are of type \mintinline{haskell}{ByteString}, since they are all
just binary strings in Redis. If a user wants to store values of
arbitrary types, he or she will have to encode and decode them as
 \mintinline{haskell}{ByteString}s.

\subsection{Semantics errors}

Although everything in Redis is binary string, they are treated diffenrently.
Redis supports many different kind of data structures, such as strings, hashes,
 lists, etc. These values are of different data types, and some commands only
 work on certain type of values.

\paragraph{Problem 1} Consider the following example. \mintinline{text}{SET} is a command that only
works on strings. The key \mintinline{text}{some-string} is now related with a
string. If we treat it as a set and add an element to it with
\mintinline{text}{SADD}, a runtime error arises.

\begin{minted}{text}
redis> SET some-string foo
OK
redis> SADD some-string bar
(error) WRONGTYPE Operation against a key
 holding the wrong kind of value
\end{minted}

\paragraph{Problem 2} Even worse, not all strings are equal! \mintinline{text}{INCR} parses the
string value as an integer, increments it by one, and store it back as a string.
If a string value can't be parse as an integer, another runtime error arises

\begin{minted}{text}
redis> SET some-string foo
OK
redis> INCR some-string
(error) ERR value is not an integer or out
 of range
\end{minted}

\paragraph{In Hedis} The same goes for Hedis, since it's only a simple wrapping on top of Redis's
protocol written in Haskell. No semantic checking is done, neither
dynamically nor statically.

\begin{minted}{haskell}
program :: Redis (Either Reply Integer)
program = do
    set "some-string" "foo"
    sadd "some-string" ["a"]
\end{minted}

Which yields the same error as in Redis's client.

\begin{minted}{haskell}
Left (Error "WRONGTYPE Operation against a
 key holding the wrong kind of value")
\end{minted}

\subsection{Hedis as an embedded DSL}

Haskell makes it easy to build and use Domain Specific Languages (DSLs).
Hedis can be regardred as one DSL that embeds in Haskell, that has variable
bindings (between keys and values), but with very little or none semantic
checkings.

We started out from making Hedis a dynamically checked embedded DSL, with a
runtime type checker that keeps track of everything. But then we found that,
things can be a lot easier, by laveraging the host language's type checker,
and make it statically checked!

\subsection{Contributions}

To summerize our contributions:

\begin{itemize}[noitemsep]
\item We model variable bindings of an embedded DSL with language extensions
like type-level literals and data kinds.
\item We provide (yet another) an example of how to encode effects and
constraints of an action in types, with indexed monad and other language
 extensions such as closed type-families and constraints kinds.
\item Edis, a package we built for programmers. This package helps programmers
to write more reliable Redis codes, and makes Redis polymorphic by converting
back and forth from values of any types and boring ByteStrings.
\end{itemize}

\section{Implementation}
\subsection{Model bindings with Type-level Dictionaries}
\paragraph{Type-level literals}
\subsection{Bookkeeping with Type families}
\subsection{Capture effects with Parameterised Monad}
\paragraph{Type-level literals}

%
% Values have types, but keys don't. In other words, if we treat keys like
% variables, then Redis behaves likes a dynamically and strongly typed language.

% We can only reuse this key as a \emph{set} after it is removed with \mintinline{text}{DEL}
% explicitly.
%
% \begin{minted}{text}
% redis> DEL some-string
% (integer) 1
% redis> SADD some-string bar
% (integer) 1
% \end{minted}
